\documentclass[12pt]{article}
\usepackage{amsfonts, epsfig}
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amsmath}
\pagestyle{fancy}
\lfoot{\texttt{coms10013.github.io}}
\lhead{Analysis - 2 Extrema in more than one dimesion - Conor}
\rhead{\thepage}
\cfoot{}
\begin{document}

\section*{Extrema}

Just as in one dimension we define a critical point as $df/dx=0$ and
all extrema are critical points; in higher dimensions a critical point
is one where all the partial derivatives are zero. We saw that
\begin{equation}
  \nabla_{\mathbf{w}}f=\mathbf{w}\cdot\nabla{}f
\end{equation}
and so, given that the rate of change at an extremum must be zero in
all directions, all extrema are critical points in higher dimensions
too.

Distinguishing between maxima and minima is a small bit more
complicated in higher dimesions; it relies on the \textsl{Hessian},
the matrix of second-order partial derivatives. Here we will look at
two dimensions as an example, though the idea is similar for other
numbers of dimensions. In two dimensions the Hessian is
\begin{equation}
  H=\left(\begin{array}{cc}\frac{\partial^2f}{\partial x^2}&\frac{\partial^2f}{\partial x\partial y}\\
    \frac{\partial^2f}{\partial y\partial x}&\frac{\partial^2f}{\partial y^2}\end{array}\right)
\end{equation}
or, using the obvious notation
\begin{equation}
  H=\left(\begin{array}{cc}f_{xx}&f_{xy}\\f_{yx}&f_{yy}\end{array}\right)
\end{equation}
Also, note that for all but very weird functions $f_{xy}=f_{yx}$.

Now imagine the Hessian at a critical point looks like
\begin{equation}
  H=\left(\begin{array}{cc}-1&0\\0&-1\end{array}\right)
\end{equation}
then it would seem that we have a maximum since the $x$ and $y$
derivatives are falling at this point, just like the one-dimensional
case. More generally though there are non-zero $f_{xy}$ so what is
important is not the diagonal elements but the eigenvalues. If both
eigenvalues of the Hessian are negative at a critical point, then it's
a maximum, if both are positive it is a minimum. There is another
case, where one is positive and one is negative, that implies the
critical point is a minumum in one direction and a maximum in another;
that can happen in two dimensions and corresponds to a
\textsl{saddlepoint}, like the point where two hills join and so the
land goes up in one set of directions but down in another. Of course,
there is always the possibility of a zero eigenvalue, this is like the
one-dimensional case where the second derivative is zero, we could
still have an extremum, but a very flat one, or it might not be an
extremum at all.

This all seems to imply you need the eigenvalues of the Hessian to
classify two-dimensional critical points and certainly if you have the
eigenvalues and you can do the classification, however, you actually
only need the signs of the two eigenvalues and you can usually work
that out without actually knowing the eigenvalues. If $\lambda_1$ and
$\lambda_2$ are the eigenvalues of a matrix $H$ then
\begin{equation}
  \det{H}=\lambda_1\lambda_2
\end{equation}
and
\begin{equation}
  \rm{tr}{H}=\lambda_1+\lambda_2
\end{equation}
There are known as matrix invariants. It means that if the determinant
is negative you have a saddlepoint, if it is positive then the two
eigenvalues have the same sign, so if the trace is negative you have a
maximum, if it is negative, a minimum. If the determinant is zero then
there is a zero eigenvalue and the Hessian alone isn't enough to tell
you what is happening.

Lets do an example:
\begin{equation}
  f(x,y)=x^3+x^2y-y^2-4y
\end{equation}
Now $f_x=3x^2+2xy$ and $f_y=x^2-2y-4$. These equations are often a
pain, it is hard to come up with good examples were the equations for
the critical points are easy to solve. In this example, the first equation looks promising:
\begin{equation}
  3x^2+2xy=0\implies{}x(3x+2y)=0
\end{equation}
so the solutions are $x=0$ or $x=-2y/3$. If $x=0$ then the second
equation gives $y=-2$; if $x=-2y/3$ the second equation is
\begin{equation}
  4y^2/9-2y-4=0\implies{}2y^2-9y-18=0
\end{equation}
and this factorizes into $(2y+3)(y-6)=0$ so the solutions are $y=-3/2$
and $y=6$. Hence the three critical points are $(0,-2)$, $(-4,6)$ and $(1,-3/2)$. Now lets work out the Hessian:
\begin{equation}
  H=\left(\begin{array}{cc}6x+2y&2x\\2x&-2\end{array}\right)
\end{equation}
so for the point $(0,-2)$
\begin{equation}
  H=\left(\begin{array}{cc}-4&0\\0&-2\end{array}\right)
\end{equation}
and since this is diagonal matrix we can read off the eigenvalues,
they are $-4$ and $-2$ and so this is a maximum. For $(-4,6)$ the Hessian is
\begin{equation}
  H=\left(\begin{array}{cc}-12&-8\\-8&-2\end{array}\right)
\end{equation}
and hence the determinant of the Hessian is $-40$ which is negative and there is a saddle point. For $(1,-3/2)$ the Hessian is
\begin{equation}
  H=\left(\begin{array}{cc}3&2\\2&-2\end{array}\right)
\end{equation}
so the determinant is -10 and this is another saddle point.

For higher dimensions the story is much the same, but now there are
more types of saddle point, for example in three dimensions there are
saddle points with two up directions and one down and saddle points
with one up direction and two down. These will correspond to two
negative and one positive eigenvalues, or one negative and two
positive eigenvalues for the Hessian.

\section*{Summary}

In higher dimensions there is a critical point when the gradient is zero. To classify the critical point the Hessian is calculated, in two-dimension this is
  \begin{equation}
  H=\left(\begin{array}{cc}f_{xx}&f_{xy}\\f_{yx}&f_{yy}\end{array}\right)
\end{equation}
If this has two positive eigenvalues at a critical point then it is a
minimum, two negative, a maximum, one of each, a saddle point. If one,
or both, of the eigenvalues is zero then it could be any of the three,
or it might not be an extremum at all. The sign of the eigenvalues can
be figured out from the determinant and trace since the determinant is
the product of the eigenvalues and the trace is the sum.

  
\end{document}

